---
title: "Data Manipulation with dplyr"
author: "Chris DeAngelis, CFA"
date: "December 3, 2017"
output: html_document
---

Adapted from my coursework on Datacamp.
https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial

The first step is to load the required packages. 
```{r setup, include=TRUE}
# Load Packages
library(dplyr) # Tools for manipulating data frames
library(hflights) # Sample dataset on flihts departing Houston airports
library(RMySQL) # MySQL driver 
```

In chapter 1, I preview the data and formatting.
```{r chapter1}
## Chapter 1 - Introduction to dplyr and tbls
# Call both head() and summary() on hflights
head(hflights)
summary(hflights)
# Convert the hflights data.frame into a hflights tbl
hflights <- as.tbl(hflights)

# Display the hflights tbl
hflights

# Create the object carriers
carriers <- hflights$UniqueCarrier

# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add the Carrier column to hflights
#Carrier <- hflights$UniqueCarrier
hflights$Carrier <- lut[hflights$UniqueCarrier]

# Glimpse at hflights
glimpse(hflights)

# The lookup table
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", "D" = "security", "E" = "not cancelled")

# Add the Code column
hflights$Code <- lut[hflights$CancellationCode]

# Glimpse at hflights
glimpse(hflights)
```

In chapter 2, I slice the data in a few different ways and calculate time spent taxiing and on the ground.
```{r chapter2}
## Chapter 2 - Select and mutate
# Print out a tbl with the four columns of hflights related to delay
select(hflights,ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# Print out the columns Origin up to Cancelled of hflights
select(hflights, 14:19)

# Answer to last question: be concise!
select(hflights, 1:4,12:21)

# Print out a tbl containing just ArrDelay and DepDelay
select(hflights,ArrDelay,DepDelay)

# Print out a tbl as described in the second instruction, using both helper functions and variable names
select(hflights,UniqueCarrier,ends_with("Num"),starts_with("Cancel"))

# Print out a tbl as described in the third instruction, using only helper functions.
select(hflights,contains("Tim"),contains("Del"))

# Finish select call so that ex1d matches ex1r
ex1r <- hflights[c("TaxiIn", "TaxiOut", "Distance")]
ex1d <- select(hflights, contains("Taxi"),Distance)

# Finish select call so that ex2d matches ex2r
ex2r <- hflights[c("Year", "Month", "DayOfWeek", "DepTime", "ArrTime")]
ex2d <- select(hflights,1:6,-3)

# Finish select call so that ex3d matches ex3r
ex3r <- hflights[c("TailNum", "TaxiIn", "TaxiOut")]
ex3d <- select(hflights,starts_with("Tail"),contains("Taxi"))

# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)

# Add the new variable GroundTime to g1. Save the result as g2.
g2 <- mutate(g1, GroundTime = TaxiIn + TaxiOut)

# Add the new variable AverageSpeed to g2. Save the result as g3.
g3 <- mutate(g2, AverageSpeed = Distance / AirTime * 60)

# Print out g3
g3

# Add a second variable loss_ratio to the dataset: m1
m1 <- mutate(hflights, loss = ArrDelay - DepDelay, loss_ratio = loss / DepDelay)

# Add the three variables as described in the third instruction: m2
m2 <- mutate(hflights, TotalTaxi = TaxiIn + TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)
```

Chapter 3 asks questions of the data. How do JetBlue, Southwest, and Delta compare on delays out of Houston airport? How many flights flew over 3,000 miles? How many flights flew to JFK? Lastly, I sort the data by delays and cancellations.
```{r chapter3}
## Chapter 3 - Filter and arrange
# All flights that traveled 3000 miles or more
filter(hflights, Distance >= 3000)

# All flights flown by one of JetBlue, Southwest, or Delta
filter(hflights, UniqueCarrier %in% c('JetBlue', 'Southwest', 'Delta'))

# All flights where taxiing took longer than flying
filter(hflights, (TaxiIn + TaxiOut - AirTime) > 0)

# All flights that departed before 5am or arrived after 10pm
filter(hflights, DepTime < 500 | ArrTime > 2200)

# All flights that departed late but arrived ahead of schedule
filter(hflights, DepDelay > 0 & ArrDelay < 0)

# All flights that were cancelled after being delayed
filter(hflights, DepDelay > 0 & Cancelled == 1)

# Select the flights that had JFK as their destination: c1
c1 <- filter(hflights, Dest == 'JFK')

# Combine the Year, Month and DayofMonth variables to create a Date column: c2
c2 <- mutate(c1, Date = paste(Year,Month, DayofMonth, sep = "-"))

# Print out a selection of columns of c2
select(c2, Date, DepTime, ArrTime, TailNum)

# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# Arrange dtc by departure delays
arrange(dtc,DepDelay)

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode)

# Arrange dtc according to carrier and departure delays
arrange(dtc,UniqueCarrier, DepDelay)

# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))

# Arrange flights by total delay (normal order).
arrange(hflights, DepDelay + ArrDelay)
```

In chapter 4, I summarize the data by a few different statistics including delays and miles per hour. I also take a deeper look at statistics on American Airlines.
```{r chapter4}
## Chapter 4 - Summarize and the pipe operator
# Print out a summary with variables min_dist and max_dist
summarise(hflights, min_dist = min(Distance), max_dist = max(Distance))

# Print out a summary with variable max_div
summarise(filter(hflights, Diverted == 1), max_div = max(Distance))

# Remove rows that have NA ArrDelay: temp1
temp1 <- filter(hflights, !is.na(ArrDelay))

# Generate summary about ArrDelay column of temp1
summarise(temp1, earliest = min(ArrDelay), average = mean(ArrDelay), latest = max(ArrDelay), sd = sd(ArrDelay))

# Keep rows that have no NA TaxiIn and no NA TaxiOut: temp2
temp2 <- filter(hflights, (!is.na(TaxiIn) & !is.na(TaxiOut)))

# Print the maximum taxiing difference of temp2 with summarise()
summarise(temp2, max_taxi_diff = max(abs(TaxiIn - TaxiOut)))

# Generate summarizing statistics for hflights
summarise(hflights,
          n_obs = n(),
          n_carrier = n_distinct(UniqueCarrier),
          n_dest = n_distinct(Dest))

# All American Airline flights
aa <- filter(hflights, UniqueCarrier == "American")

# Generate summarizing statistics for aa 
summarise(aa,n_flights = n(), n_canc = sum(Cancelled == '1'), avg_delay = mean(ArrDelay, na.rm = TRUE))

# Write the 'piped' version of the English sentences.
hflights %>%
mutate(diff = TaxiOut - TaxiIn) %>%
filter(!is.na(diff)) %>%
summarise(avg = mean(diff))

# Chain together mutate(), filter() and summarise()
hflights %>%
mutate(RealTime = ActualElapsedTime + 100, mph = Distance / RealTime * 60) %>%
filter(!is.na(mph) & mph < 70) %>%
summarise(n_less = n(), n_dest = n_distinct(Dest), min_dist = min(Distance), max_dist = max(Distance))

# Finish the command with a filter() and summarise() call
hflights %>%
  mutate(RealTime = ActualElapsedTime + 100, mph = Distance / RealTime * 60) %>%
  filter(mph < 105 | Cancelled == '1' | Diverted == '1') %>%
  summarise(n_non = n(), n_dest = n_distinct(Dest), min_dist = min(Distance), max_dist = max(Distance))
  
# Count the number of overnight flights
hflights %>%
filter(!is.na(DepTime) & !is.na(ArrTime) & DepTime > ArrTime) %>%
summarise(num = n())
```

Chapter 5 summarizes the data to answer questions like: who has the most cancelled flights? Who has the longest average delay? How many airplanes only flew to one location. I also have an introduction into connecting to and extracting from a database.
```{r chapter5}
## Chapter 5 - Group_by and working with databases
# Make an ordered per-carrier summary of hflights
hflights %>%
  group_by(UniqueCarrier) %>%
  summarise(p_canc = mean(Cancelled == 1)*100,
            avg_delay = mean(ArrDelay, na.rm = TRUE)) %>%
  arrange(avg_delay, p_canc)

# Ordered overview of average arrival delays per carrier
hflights %>%
  filter(!is.na(ArrDelay) & ArrDelay > 0) %>%
  group_by(UniqueCarrier) %>%
  summarise(avg = mean(ArrDelay)) %>%
  mutate(rank = rank(avg)) %>%
  arrange(rank)

# How many airplanes only flew to one destination?
hflights %>%
  group_by(TailNum) %>%
  summarise(ndest = n_distinct(Dest)) %>%
  filter(ndest == 1) %>%
  summarise(nplanes = n())

# Find the most visited destination for each carrier
hflights %>% 
  group_by(UniqueCarrier, Dest) %>%
  summarise(n = n()) %>%
  mutate(rank = rank(desc(n))) %>%
  filter(rank == 1)

# Use summarise to calculate n_carrier
summarise(hflights, n_carrier = n_distinct(UniqueCarrier))
  
# Set up a connection to the mysql database
my_db <- src_mysql(dbname = "dplyr", 
                   host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                   port = 3306, 
                   user = "student",
                   password = "datacamp")

# Reference a table within that source: nycflights
nycflights <- tbl(my_db, "dplyr")

# glimpse at nycflights
glimpse(nycflights)

# Ordered, grouped summary of nycflights
  nycflights %>%
    group_by(carrier) %>%
    summarise(n_flights = n(), avg_delay = mean(arr_delay)) %>%
    arrange(avg_delay)

```

